---
title: "ADMIXTURE Global Ancestry Inference"
author: "Brian Fulton-Howard"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(ggplot2.discrete.fill = pals::cols25(25))

suppressPackageStartupMessages(library(dplyr))
library(readr)
library(tidyr)
library(purrr)
library(tibble)
library(stringr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(ggforce)
library(ggrepel)

projpaths <- sf_proj_search_paths() |>
  (\(x) x[dir.exists(x)])() |>
  sf_proj_search_paths()

if (exists("snakemake")) {
  load(snakemake@input[[1]])
} else {
  load("../../reference_proc/hgdp_1kg.admixture.rda")
}
```

# Sample: `r snakemake@wildcards$cohort`

Build: `r snakemake@wildcards$build`

## Reference Processing

To prepare the gnomAD reference, we did the following:

*  Remove samples without a population inference from gnomAD or without high_quality set to TRUE.
*  Make a column (`spop`) by doing the following with the populations inferred by gnomAD:
   *  Merge "nfe" and "fin" into "EUR"
   *  Move oceanic subjects from "oth" to their own "OCE" category.
   *  Capitalize all other superpopulations.
*  Make a column (`spop_checked`) where the original superpopulations match the inferred superpopulations:
   *  The new `spop` column is used for inferred superpopulation.
   *  The `genetic_region` column is used for original superpopulation.
   *  "CSA" in `genetic_region` is considered a match to "SAS" in `spop`. "SAS" is used in the new column. 
   *  All subjects where there is no match are set to "NA"

## ADMIXTURE Procedure

The following steps are used to generate Global Ancestry Inference (GAI) estimates:

1. Process the reference label data as described above.
2. Obtain the intersection of the reference and target varients, then prune the reference with a 100kb window and R^2 of 0.1.
3. Restrict sample genotypes to those present in the pruned reference, then merge with the reference samples. Check that the `.bim` files are identical.
4. Run unsupervised ADMIXTURE with K = 12 on the reference dataset.
5. Run ADMIXTURE projection on the merged reference and target samples.
6. Read in the processed reference labels, ADMIXTURE cluster estimates (Q files), and PLINK `.fam` files.
7. Merge the reference labels with the ADMIXTURE cluster estimates and extract the reference samples for labeling, excluding Middle Eastern reference samples.
8. Label the clusters by assigning to each cluster the superpopulation with the highest average proportion within that cluster. The checked superpopulation labels are used for this labeling process.
9. Using the cluster labels, calculate GAI proportions and maximum superpopulation for all samples.
10. Visualize below.

## Cluster label assignments:

```{r label_clusters, echo=FALSE, fig.height=9, fig.width=7, cache=FALSE}
heatmap(heatmap_mat)
```

```{r prep_plotting, include=FALSE}
plot_admix <- function(x, by = cluster_cols, poplabel = TRUE) {
  stripangle <- 90
  tbl_long <- x |>
    pivot_longer(all_of(by), names_to = "Cluster", values_to = "prop" )

  if (all(str_detect(by, "^k\\d+$"))) {
    tbl_long <- mutate(tbl_long, Cluster = assign_cname_vec[Cluster])
  }

  if ("genetic_region" %in% names(x)) {
    tbl_use <- tbl_long |>
      mutate(spop = ifelse(genetic_region == "CSA", "SAS", genetic_region))
  } else {
    tbl_use <- tbl_long
  }

  if (!poplabel) {
    pop_number <- x |>
      group_by(spop) |>
      mutate(label_n = sprintf("%s\n%i", spop, n())) |>
      ungroup() |>
      select(ID, label_n)

    tbl_use <- tbl_use |>
      arrange(admixture_super_pop_max, -max_spop_prop) |>
      left_join(pop_number, by = "ID") |>
      mutate(spop = label_n,
             ID = forcats::fct_inorder(as.character(ID)))
    stripangle <- 0
  }

  plt <- ggplot(tbl_use, aes(x = ID, y = prop, fill = Cluster)) +
    geom_bar(position = "fill", stat = "identity", width = 1) +
    theme_classic() +
    labs(x = "Individual", y = "Global Ancestry", color = "Cluster") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.y = element_blank(),
          axis.title.x = element_blank(),
          panel.grid.major.x = element_blank(),
          strip.text.x = element_text(angle = stripangle))
  if (poplabel) {
    plt + ggh4x::facet_nested(~ spop + pop, switch = "x",
                              scales = "free", space = "free")
  } else if (!all(is.na(x$spop))) {
    plt + facet_grid(~ spop, switch = "x",
                     scales = "free", space = "free")
  } else {
    plt + facet_grid(~ admixture_super_pop_max, switch = "x",
                     scales = "free", space = "free")
  }
}

plot_map <- function(df_pop, seed, force_push = 1000000, force_pull = 10) {
  findboxes <- function(df, xcol, ycol, size_x, size_y,
                        xlimit = c(-Inf, Inf), ylimit = c(-Inf, Inf),
                        force_push = 1, force_pull = 1, maxiter = 200000) {

    boxdf <- tibble(x1 = df[[xcol]] - (size_x / 2),
                    y1 = df[[ycol]] - (size_y / 2),
                    x2 = df[[xcol]] + (size_x / 2),
                    y2 = df[[ycol]] + (size_y / 2))

    boxmatrix <- as.matrix(boxdf)

    pointmatrix <- df[c(xcol, ycol)] |>
      as.matrix()

    ggrepel:::repel_boxes2(data_points = pointmatrix,
                           point_padding_x = 0.25,
                           point_padding_y = 0,
                           boxes = boxmatrix,
                           xlim = xlimit,
                           ylim = ylimit,
                           force_push = force_push * 1e-6,
                           force_pull = force_pull * 1e-2,
                           max_iter = maxiter,
                           point_size = rep_along(df[[xcol]], 0.1),
                           hjust = rep_along(df[[xcol]], 0.5),
                           vjust = rep_along(df[[xcol]], 0.5),
                           max_overlaps = Inf) |>
      rename(x_moved = x, y_moved = y) |>
      bind_cols(boxdf) |>
      add_column(df, .before = 1)
  }

  reduce_end <- function(x1, y1, x2, y2, r) {
    x <- x2 - x1
    y <- y2 - y1
    mag <- sqrt(x ^ 2 + y ^ 2)
    list(
      x = x2 - r * x / mag,
      y = y2 - r * y / mag
    )
  }

  pie_radius <- function(n) log(n)
  pie_radius_inner <- function(n) log(n) / 2.5

  set.seed(seed)

  mapdata_raw <-
    df_pop |>
    group_by(pop) |>
    summarise(across(matches("^k\\d+$"), mean),
              across(all_of(c("latitude", "longitude")),
                     \(x) mean(as.numeric(x))),
              n = n()) |>
    rename(all_of(assign_cname_vec_i))

  mapdata <- mapdata_raw |>
    findboxes("longitude", "latitude",
              log(mapdata_raw$n) * 2, log(mapdata_raw$n) * 2,
              c(-180, 180), c(-90, 90), 1000000, 10) |>
    mutate(lineend = transpose(reduce_end(
      longitude, latitude, x_moved, y_moved, pie_radius_inner(n)))) |>
    tidyr::unnest_wider(col = lineend, names_sep = "_")

  mapdata |>
    ggplot() +
    geom_sf(data = world, fill = "gray92", color = "gray80") +
    geom_link(aes(x = longitude, y = latitude,
                  xend = lineend_x, yend = lineend_y),
              arrow = arrow(angle = 30, length = unit(3, "points"),
                            ends = "first", type = "open")) +
    geom_arc_bar(
      data = \(x) {
        pivot_longer(x, cols = all_of(unname(assign_cname_vec)),
                     names_to = "cluster", values_to =  "amount")
      },
      aes(x0 = x_moved, y0 = y_moved, r0 = pie_radius_inner(n),
          r = pie_radius(n), amount = amount, fill = cluster),
      stat = "pie", color = NA) +
    geom_circle(aes(x0 = x_moved, y0 = y_moved, r = pie_radius(n)),
                fill = NA, color = "black", linewidth = 0.2) +
    geom_circle(aes(x0 = x_moved, y0 = y_moved, r = pie_radius_inner(n)),
                fill = NA, color = "black", linewidth = 0.2) +
    theme_classic() +
    coord_sf(expand = FALSE) +
    ggtitle("Map of average cluster proportions in each population") + 
    xlab("Longitude") + ylab("Latitude")
}
```

```{r filter_ref, include=FALSE}
admix_ref <- filter(out_admix, partition != "sample")
```

## Reference plots

Below is a plot of the cluster proportions for all individuals who were assigned a superpopulation in gnomAD. The original superpopulation and population labels are used here.

```{r cplot_unfilt_pop1, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_ref |>
  filter(genetic_region %in% c("AFR", "AMR", "EUR")) |>
  plot_admix() + ggtitle("Admixture Plot by Cluster (AFR, AMR and EUR)")
```

```{r cplot_unfilt_pop2, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_ref |>
  filter(!(genetic_region %in% c("AFR", "AMR", "EUR"))) |>
  plot_admix() + ggtitle("Admixture Plot by Cluster (not AFR, AMR or EUR)")
```

This is a map of average cluster proportions in each population:

```{r map, echo=FALSE, fig.width = 14, fig.height = 7.5, warning = FALSE, cache = TRUE}
world <- ne_countries(scale = "medium", returnclass = "sf")

plt <- plot_map(admix_ref, 1007)

plt
```

For simplified viewing, here is a cluster plot broken down only by superpopulation and with summed superpopulation clusters:

```{r cplot_unfilt_spop, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
filter_output <- \(tbl, max_prop) {
  filter(tbl, !is.na(spop_checked) &
           spop_checked == as.character(admixture_super_pop_max) &
           spop_checked != "OCE" &
           max_spop_prop >= max_prop)
}

admix_ref |>
  filter_output(0) |>
  plot_admix(by = superpops, poplabel = FALSE) +
  ggtitle("Admixture Plot by Continental Ancestry (no admixture filter)")
```

Only individuals for whom gnomAD population matches ADBIXTURE population are shown, and Oceanic references are removed.

## Target

```{r filter_noref, include=FALSE}
admix_noref <- filter(out_admix, partition == "sample")
```

```{r cplot_samp_pop1, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_noref |>
  filter(admixture_super_pop_max %in% c("AFR", "AMR", "EUR")) |>
  plot_admix(poplabel = FALSE) +
    ggtitle("Admixture Plot by Cluster (AFR, AMR and EUR)")
```

```{r cplot_samp_pop2, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_noref |>
  filter(!(admixture_super_pop_max %in% c("AFR", "AMR", "EUR"))) |>
  plot_admix(poplabel = FALSE) +
    ggtitle("Admixture Plot by Cluster (not AFR, AMR or EUR)")
```

Below is a plot with varying maximum ADMIXTURE superpopulation proportion filters. After filtering, the following sample sizes remain:

```{r filt_sampsize, echo=FALSE, fig.width = 9, fig.height = 7, warning = FALSE, cache = TRUE}
c(0.75, 0.8, 0.85, 0.9, 0.95) |>
  map_dfr(\(prop) {
    filter(admix_noref, max_spop_prop >= prop) |>
      count(admixture_super_pop_max) |>
      mutate("Proportion filter" = prop)
  }) |>
  rename(Superpopulation = admixture_super_pop_max) |>
  ggplot() + theme_bw() +
  geom_line(aes(x = `Proportion filter`, y = n, color = Superpopulation))
```

Here are the cluster plots at various proportion filters:

```{r cplot_filt_80, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_noref |>
  filter(max_spop_prop >= 0.8) |>
  plot_admix(poplabel = FALSE) +
  ggtitle("Admixture Plot by Continental Ancestry (admixture filter >= 0.8)")
```

```{r cplot_filt_90, echo=FALSE, fig.height=7, fig.width=16, message=FALSE, warning=FALSE}
admix_noref |>
  filter(max_spop_prop >= 0.9) |>
  plot_admix(poplabel = FALSE) +
  ggtitle("Admixture Plot by Continental Ancestry (admixture filter >= 0.9)")
```

